---
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Probability and Statistics

# Lab Assignment 1: Naive Bayes Classifier

## Work breakdown

-   *Name1 Surname1*: Mykhailo Rykhalskyi
-   *Name2 Surname2*: Mariia Hamaniuk
-   *Name3 Surname3*: Ivan Maksymchuk

## Introduction

During the first three weeks, you learned a couple of essential notions
and theorems, and one of the most important among them is the **Bayes
theorem**.

**Naive Bayes Classifier** is a simple algorithm, which is based on
**Bayes theorem** and used for solving classification problems.
**Classification problem** is a problem in which an observation has to
be classified in one of the $n$ classes based on its similarity with
observations in each class.

It is a **probabilistic classifier**, which means it predicts based on
the probability of an observation belonging to each class. To compute
it, this algorithm uses **Bayes' formula,** that you probably already
came across in **Lesson 3:**
$$\mathsf{P}(\mathrm{class}\mid \mathrm{observation})=\frac{\mathsf{P}(\mathrm{observation}\mid\mathrm{class})\mathsf{P}(\mathrm{class})}{\mathsf{P}(\mathrm{observation})}$$

Under the strong **independence** assumption, one can calculate
$\mathsf{P}(\mathrm{observation} \mid \mathrm{class})$ as
$$\mathsf{P}(\mathrm{observation}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i), \qquad \mathsf{P}(\mathrm{observation} \mid \mathrm{class}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i \mid \mathrm{class}),$$
where $n$ is the total number of features describing a given observation
(*For example, if an observation is presented as a sentence, then each
word can be a feature*). Thus,
$\mathsf{P}(\mathrm{class}|\mathrm{observation})$ now can be calculated
as

$$\mathsf{P}(\mathrm{class} \mid \mathrm{\mathrm{observation}}) = \mathsf{P}(\mathrm{class})\times \prod_{i=1}^{n}\frac{\mathsf{P}(\mathrm{feature}_i\mid \mathrm{class})}{\mathsf{P}(\mathrm{feature}_i)}\tag{1}$$

All the terms on the right-hand side can be estimated as corresponding
relative frequencies using available data\

**See [*this
link*](https://www.javatpoint.com/machine-learning-naive-bayes-classifier)
for more detailed explanations & examples :) Also you can watch [*this
video*](https://youtu.be/O2L2Uv9pdDA?si=-ohkHVDuu3sLLGMq) for more
examples!**

## Data description

There are 5 datasets uploaded on the cms (data.zip)

To determine your variant, take your team number from the list of teams
on cms and take *mod 5* - this is the number of your data set.

-   **0 - authors** This data set consists of citations of three famous
    writers: Edgar Alan Poe, Mary Wollstonecraft Shelley and HP
    Lovecraft. The task with this data set is to classify a piece of
    text with the author who was more likely to write it.

-   ***1 - discrimination*** **This data set consists of tweets that
    have discriminatory (sexism or racism) messages or of tweets that
    are of neutral mood. The task is to determine whether a given tweet
    has discriminatory mood or does not.**

-   **2 - fake news** This data set contains data of American news: a
    headline and an abstract of the article. Each piece of news is
    classified as fake or credible. The task is to classify the news
    from test.csv as credible or fake.

-   **3 - sentiment** All the text messages contained in this data set
    are labeled with three sentiments: positive, neutral or negative.
    The task is to classify some text message as the one of positive
    mood, negative or neutral.

-   **4 - spam** This last data set contains SMS messages classified as
    spam or non-spam (ham in the data set). The task is to determine
    whether a given message is spam or non-spam.

Each data set consists of two files: *train.csv* and *test.csv*. The
first one is used to find the probabilities of the corresponding classes
and the second one is used to test your classifier afterwards. Note that
it is crucial to randomly split your data into training and testing
parts to test the classifierʼs possibilities on the unseen data.

```{r}
# here goes a list of recommended libraries,
# though you may install other ones if they are needed
library(tidytext)
library(readr)
library(dplyr)
library(ggplot2)
library(tm)
library(tibble)
library(stringr)
```

## Outline of the work

1.  **Data pre-processing** (includes removing punctuation marks and
    stop words, representing each message as a bag-of-words)
2.  **Data visualization** (it's time to plot your data!)
3.  **Classifier implementation** (using the training set, calculate all
    the conditional probabilities in formula (1))
4.  **Measurements of effectiveness of your classifier** (use the
    results from the previous step to predict classes for messages in
    the testing set and measure the accuracy, precision and recall, F1
    score metric etc)
5.  **Conclusions**

*!! do not forget to submit both the (compiled) Rmd source file and the
.html output !!*

## Data pre-processing

-   Read the *.csv* data files.
-   Сlear your data from punctuation or other unneeded symbols.
-   Clear you data from stop words. You don't want words as is, and, or
    etc. to affect your probabilities distributions, so it is a wise
    decision to get rid of them. Find list of stop words in the cms
    under the lab task.
-   Represent each test message as its bag-of-words. Here:
    <https://machinelearningmastery.com/gentle-introduction-bag-words-model/>
    you can find general introduction to the bag-of-words model and
    examples on to create it.
-   It is highly recommended to get familiar with R dataframes, it would
    make the work much easier to do.
-   Useful links:
    -   <https://steviep42.github.io/webscraping/book/bagofwords.html#tidytext> -
        example of using *tidytext* to count frequencies of the words.
    -   Basics of Text Mining in R:
        <http://rstudio-pubs-static.s3.amazonaws.com/256588_57b585da6c054349825cba46685d8464.html>
        . Note that it also includes an example on how to create a bag
        of words from your text document.

```{r}
list.files(getwd())
list.files("1-discrimination")
```

```{r}
test_path <- "data/test.csv"
train_path <- "data/train.csv"

stop_words <- read_file("data/stop_words.txt")
# https://stackoverflow.com/questions/27195912/why-does-strsplit-return-a-list
splitted_stop_words <- strsplit(stop_words, split='\n')
splitted_stop_words <- splitted_stop_words[[1]]
```

```{r}
train <-  read.csv(file = train_path, stringsAsFactors = FALSE)
test <-  read.csv(file = test_path, stringsAsFactors = FALSE)
```

```{r}
# note the power functional features of R bring us! 
# tidy_text <- unnest_tokens(train, 'splitted', 'text', token="words") %>%
             #filter(!splitted %in% stop_words)

# tidy_text %>% count(splitted,sort=TRUE)
             
create_bow_dataframe <- function(file, stop_words_list) {
  
  
  stop_tbl <- tibble(word = unique(tolower(as.character(stop_words_list))))
  
  tidy_text <- file %>%
    mutate(doc_id = row_number(),
           tweet  = str_replace_all(str_to_lower(tweet), "[^a-z]+", " ")) %>%
    unnest_tokens(word, tweet, to_lower = FALSE) %>%
    anti_join(stop_tbl, by = "word") %>%
    right_join(tibble(doc_id = seq_len(nrow(file))), by = "doc_id") %>%
    arrange(doc_id)
  
  # 3. Підрахунок частоти слів
  word_counts <- tidy_text %>%
    count(doc_id, word, sort = FALSE)
  
  # 4.1 Створення Document-Term Matrix (DTM) за допомогою tidytext::cast_dtm
  # Ця функція замінює собою весь блок pivot_wider.
  # Результат: Розріджена матриця, де Рядки = Документи, Стовпці = Терміни.
  dtm_sparse <- word_counts %>%
    cast_dtm(document = doc_id, term = word, value = n)
  
  # 4.2 Перетворення розрідженої матриці (DTM) на щільну (Dense Matrix)
  # Це потрібно, оскільки багато ML-алгоритмів вимагають щільну матрицю.
  bag_of_words <- as.matrix(dtm_sparse)
  
  # 5. Вектор міток (0/1)
  labels <- ifelse(file$label == "discrim", 1, 0)
  head(bag_of_words)
  return(list(bag_of_words = bag_of_words, labels = labels))
}

train <- create_bow_dataframe(train, splitted_stop_words)

X_train = train$bag_of_words
y_train = train$labels


test <- create_bow_dataframe(test, splitted_stop_words)

X_test = test$bag_of_words
y_test = test$labels
```

## Data visualization

Each time you work with some data, you need to understand it before you
start processing it. R has very powerful tools to make nice plots and
visualization. Show what are the most common words for negative and
positive examples as a histogram, word cloud etc. Be creative!

```{r}
# Load necessary libraries (make sure these are installed: install.packages(c("tidyverse", "tidytext")))
library(tidyverse)
library(tidytext)
library(stringr)

# --- Data Preparation Function ---

get_word_counts <- function(file, stop_words_list) {
  
  # Create a tibble of stop words for anti_join
  stop_tbl <- tibble(word = unique(tolower(as.character(stop_words_list))))
  
  # --- 1. Tidy the text (Keeping 'label' throughout) ---
  tidy_text <- file %>%
    # Create doc_id for consistency (but it's not strictly needed here anymore)
    mutate(doc_id = row_number(),
           # Convert to lower case, remove non-alphabetic characters (except spaces)
           tweet = str_replace_all(str_to_lower(tweet), "[^a-z]+", " ")) %>%
    # 'label' is kept automatically by the pipe
    # Unnest tokens (this now keeps doc_id and label alongside 'word')
    unnest_tokens(word, tweet, to_lower = FALSE) %>%
    # Remove stop words
    anti_join(stop_tbl, by = "word") %>%
    # Keep only words with at least 2 characters
    filter(str_length(word) >= 2) %>%
    # Keep only the columns needed for counting
    select(label, word)
  
  # --- 2. Count the words by label ---
  word_counts <- tidy_text %>%
    count(label, word, sort = TRUE) %>%
    # Group by label to calculate the relative frequency
    group_by(label) %>%
    mutate(total_words = sum(n),
           proportion = n / total_words) %>%
    ungroup()
  
  return(word_counts)
}

# --- Apply the function to the 'train' dataset ---

# ---
train_raw <- read.csv(file=train_path, stringsAsFactors = FALSE)
# Get the processed word count data

# *** MODIFIED LINE: Take only the first 1000 rows to save memory ***
train_subset <- train_raw %>% 
  slice(1:1000)

word_freqs <- get_word_counts(train_subset, splitted_stop_words)


# --- Prepare Data for Butterfly Chart ---

# 1. Filter for the top N words overall (e.g., top 15 words in BOTH categories)
top_n <- 15 
# Find the top 15 words by their total count across both labels
top_words <- word_freqs %>%
  group_by(word) %>%
  summarise(total_n = sum(n)) %>%
  slice_max(total_n, n = top_n * 2) %>% # Get more than N to ensure coverage
  pull(word)

# 2. Reshape and modify for the chart
plot_data <- word_freqs %>%
  # Filter to only include the top words
  filter(word %in% top_words) %>%
  # Pivot wider to get 'discrim' and 'neutral' proportions side-by-side
  select(label, word, proportion) %>%
  pivot_wider(names_from = label, values_from = proportion, values_fill = 0) %>%
  # Calculate the difference for sorting
  mutate(diff = discrim - neutral) %>%
  # Sort by the difference to highlight unique words for each group
  arrange(diff) %>%
  # Convert 'word' to a factor to enforce the sorted order in the plot
  mutate(word = factor(word, levels = word)) %>%
  # Reshape back to long format for ggplot (using the factor levels)
  pivot_longer(cols = c("discrim", "neutral"), names_to = "label", values_to = "proportion") %>%
  # For the 'discrim' side, negate the proportion to make the bars point left
  mutate(percent = proportion * 100) %>%
  mutate(proportion = ifelse(label == "discrim", -percent, percent))

# --- Create the Butterfly Chart (Bilateral Bar Chart) ---

butterfly_chart <- plot_data %>%
  ggplot(aes(x = proportion, y = word, fill = label)) +
  # Use geom_bar for the bars
  geom_bar(stat = "identity", position = "identity") +
  # Add a vertical line at 0
  geom_vline(xintercept = 0, color = "black", linetype = "solid", linewidth = 0.5) +
  # Customize colors
  scale_fill_manual(values = c("discrim" = "#E37070", "neutral" = "#70A6E3"),
                    labels = c("discrim" = "Discriminatory Tweets", 
                               "neutral" = "Neutral Tweets")) +
  # Customize labels and title
  labs(title = paste("Top", length(unique(plot_data$word)), "Words: Discriminatory vs. Neutral Tweets"),
       subtitle = "Word Proportions in Each Category",
       x = "Proportion of Words (Neutral -> Positive | Discriminatory -> Negative)",
       y = NULL, # Remove y-axis label since the words are the labels
       fill = "Tweet Category") +
  # Use a theme and make it look clean
  theme_minimal() +
  theme(
    # Center and bold the title
    plot.title = element_text(hjust = 0.5, face = "bold"),
    # Customize axis text appearance
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    # Move the legend to the top
    legend.position = "top",
    # Remove grid lines for a cleaner look
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  # Use a custom scale for x-axis to show positive numbers on both sides
  scale_x_continuous(labels = abs) 

# Print the chart
print(butterfly_chart)
```

## Classifier implementation

```{r}
naiveBayes <- setRefClass("naiveBayes",
                          
       # here it would be wise to have some vars to store intermediate result
       # frequency dict etc. Though pay attention to bag of words! 
       fields = list(
            classes  = "character",
            vocab    = "character",
            priors   = "numeric",     # P(class)
            logtheta = "matrix",      # log P(term|class), rows=vocab, cols=classes ("0","1")
            alpha    = "numeric"      # Laplace smoothing
       ),
       methods = list(
                    # prepare your training data as X - bag of words for each of your
                    # messages and corresponding label for the message encoded as 0 or 1 
                    # (binary classification task)
                    fit = function(X, y)
                    {
                          if (is.data.frame(X)) X <- as.matrix(X)

                          y <- as.factor(y)

                          vocab_ <- colnames(X)

                          # Priors P(y)
                          taby <- table(y)
                          priors_ <- as.numeric(taby) / sum(taby)
                          names(priors_) <- levels(y)
                          
                          # Class-term counts with Laplace smoothing
                          a <- ifelse(is.null(.self$alpha), 1, .self$alpha)
                          counts0 <- if ("0" %in% levels(y) && taby["0"] > 0)
                            colSums(X[y == "0", , drop = FALSE]) + a else rep(a, ncol(X))
                          counts1 <- if ("1" %in% levels(y) && taby["1"] > 0)
                            colSums(X[y == "1", , drop = FALSE]) + a else rep(a, ncol(X))
                          
                          # log P(term | class)
                          logtheta_ <- cbind(
                            `0` = log(counts0 / sum(counts0)),
                            `1` = log(counts1 / sum(counts1))
                          )
                          rownames(logtheta_) <- vocab_
                          
                          # Save params
                          .self$classes  <- c("0","1")
                          .self$vocab    <- vocab_
                          .self$priors   <- priors_
                          .self$logtheta <- logtheta_
                          invisible(.self)
                    },
                    
                    # return prediction for a single message 
                    predict = function(message)
                    {
                        x  <- if (is.matrix(message)) message else matrix(message, nrow = 1)
                        if (is.null(colnames(x))) colnames(x) <- .self$vocab
                        lp <- x %*% .self$logtheta + rep(log(.self$priors), each = nrow(x))
                        as.integer(colnames(lp)[max.col(lp, ties.method = "first")])
                    },
                    
                    # score you test set so to get the understanding how well you model
                    # works.
                    # look at f1 score or precision and recall
                    # visualize them 
                    # try how well your model generalizes to real world data! 
                    score = function(X_test, y_test)
                    {

                        if (is.null(colnames(X_test))) {
                          colnames(X_test) <- .self$vocab[seq_len(min(ncol(X_test), length(.self$vocab)))]
                        }
                      
                        m <- match(.self$vocab, colnames(X_test))
                        X0 <- matrix(0, nrow(X_test), length(.self$vocab), dimnames = list(NULL, .self$vocab))
                        
                        if (any(!is.na(m))){
                          X0[, !is.na(m)] <- X_test[, m[!is.na(m)], drop = FALSE]
                        }
                        
                        X_test <- X0
                        yhat  <- apply(X_test, 1, function(row) .self$predict(row))
                        ytrue <- as.integer(as.character(y_test))
                        
                        TP <- sum(yhat==1 & ytrue==1)
                        FP <- sum(yhat==1 & ytrue==0)
                        FN <- sum(yhat==0 & ytrue==1)
                        
                        precision <- if ((TP+FP)==0) 0 else TP/(TP+FP)
                        recall    <- if ((TP+FN)==0) 0 else TP/(TP+FN)
                        f1        <- if ((precision+recall)==0) 0 else 2*precision*recall/(precision+recall)
                        
                        c(precision=precision, recall=recall, f1=f1)
                    }
))

model = naiveBayes(alpha=1)
model$fit(X_train, y_train)
model$score(X_test, y_test)
```


```{r}
#visualizing prediction, f1, and recall

# 1. Score the model to get the metrics
model_scores <- model$score(X_test, y_test)

# 2. Prepare the data for ggplot
metrics_df <- data.frame(
  Metric = names(model_scores),
  Value = as.numeric(model_scores)
)

# 3. Create the bar chart visualization
metrics_plot <- metrics_df %>%
  ggplot(aes(x = reorder(Metric, Value, desc = TRUE), y = Value, fill = Metric)) +
  geom_bar(stat = "identity", width = 0.7) +
  # Add the score value text on top of the bars
  geom_text(aes(label = round(Value, 3)), vjust = -0.5, size = 5, color = "black") +
  labs(
    title = "Naive Bayes Classifier Performance Metrics",
    subtitle = "Precision, Recall, and F1 Score",
    x = "Metric",
    y = "Score Value"
  ) +
  # Set the Y-axis limit to 0-1 for clarity
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  # Customize colors
  scale_fill_manual(values = c("precision" = "#70A6E3", "recall" = "#E37070", "f1" = "#4CAF50")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none",
    panel.grid.major.x = element_blank()
  )

print(metrics_plot)
```

## Measure effectiveness of your classifier

-   Note that accuracy is not always a good metric for your classifier.
    Look at precision and recall curves, F1 score metric.

    When evaluating the model, it's important to understand the
    different types of classification results:

    -   A ***true positive*** result is one where the model correctly
        predicts the positive class.
    -   A ***true negative*** result is one where the model correctly
        predicts the negative class.
    -   A ***false positive*** result is one where the model incorrectly
        predicts the positive class when it is actually negative.
    -   A ***false negative*** result is one where the model incorrectly
        predicts the negative class when it is actually positive.

    Precision measures the proportion of true positive predictions among
    all positive predictions made by the model.

    $$
    Precision = \frac{TP}{TP+FP}
    $$

    Recall, on the other hand, measures the proportion of true positives
    identified out of all actual positive cases.

    $$
    Recall = \frac{TP}{TP+FN}
    $$

    F1 score is the harmonic mean of both precision and recall.

    $$
    F1 = \frac{2\times Precision \times Recall}{Precision + Recall}
    $$

    **See [this
    link](https://cohere.com/blog/classification-eval-metrics) to find
    more information about metrics.**

-   Visualize them.

-   Show failure cases.

## Conclusions

Summarize your work by explaining in a few sentences the points listed
below.

-   Describe the method implemented in general. Show what are
    mathematical foundations you are basing your solution on.
-   List pros and cons of the method. This should include the
    limitations of your method, all the assumption you make about the
    nature of your data etc.
-   Explain why accuracy is not a good choice for the base metrics for
    classification tasks. Why F1 score is always preferable?
